---
title: "Code"
author: "Ben Wollack"
date: "2026/05/21"
format: pdf
editor: source
execute:
  message: false
  warning: false
---

```{r}
#| label: read-in

# read in packages
library(tidyverse)
library(here)
library(janitor)
library(lubridate)

# read in my data and store it as an object called ben-data
ben_data <- read.csv(here("data", "ben-data-tracker - Sheet1.csv"))
```

# Data Cleaning

Goals:

Clean column names

Convert dates into an ordered variable

Convert day of week to ordered variable

```{r}
#| label: data-clean

# create new object called ben_clean from ben_data
ben_clean <- ben_data |> 
  # remove capital letters, add underscores
  clean_names() |> 
  # convert dates into an ordered variable
  mutate(date_clean = ymd(date)) |> 
  # set day of the week to have levels by chronological order
  mutate(day_of_week = as_factor(day_of_week),
         day_of_week = fct_relevel(day_of_week,
                            "Monday",
                            "Tuesday",
                            "Wednesday",
                            "Thursday",
                            "Friday",
                            "Saturday",
                            "Sunday")) |> 
  # remove unneeded columns
  subset(select = -date)

# check data frame's structure to ensure columns are read correctly
str(ben_clean)
```

Day of week and date now have associated number values, so the data can be ordered.

Each figure will use different columns, so I will begin each plot with a select function.

# Figure 1 - Timeseries

Variables being used: distance walked, date, temperature, day of week

```{r}
#| label: time-clean

# create new object to only include needed columns
ben_time <- ben_clean |> 
  # select needed columns
  select(date_clean, distance_walked_km, day_of_week, high_temperature_c)
```

```{r}
#| label: timeseries

ggplot(data = ben_time,
       mapping = aes(
         x = date_clean
       )) +
  geom_line(mapping = aes(y = distance_walked_km)) +
  geom_line(mapping = aes(y = high_temperature_c))
```

Next steps: 

Scale the temperature data so that it can overlay with the distance walked data

Differentiate the colors for each, add legend

Adjust the x-axis so that display dates are Mondays

Add points, color by day of week

Rename axes

Add title

Get rid of all grid lines except the Monday ones
# Figure 2

# Figure 3
Note that this is `figure 4` in the planning document.